---
title: "EDA"
output: html_document
date: "2026-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### EDA

```{r}
library(data.table)
library(tidyverse)
library(lubridate)
library(forecast)
library(tseries)
library(corrplot)
```

```{r}
df = fread("/Users/a../Desktop/ADSP31006_TS/Project/data/ercot_processed_2018_2025.csv")
print(head(df))
print(dim(df))
```

#### Data Structure Overview

```{r}
df <- df %>% arrange(datetime)
summary(df$ERCOT_total_load_mw)
```

#### Time Series Plot

```{r}
ggplot(df, aes(x=datetime, y=ERCOT_total_load_mw)) +
  geom_line(alpha=0.6) +
  labs(title="ERCOT Total Load Over Time") +
  theme_minimal()
```
> Strong annual seasonality; increasing peak magnitude over time

#### Daily / Weekly / Monthly Pattern

```{r}
df <- df %>%
  mutate(hour = hour(datetime),
         dow = wday(datetime, label=TRUE),
         month = month(datetime, label=TRUE))

# Hourly pattern
df %>%
  group_by(hour) %>%
  summarise(avg_load = mean(ERCOT_total_load_mw)) %>%
  ggplot(aes(hour, avg_load)) +
  geom_line() +
  labs(title="Average Load by Hour")

# Weekly pattern
df %>%
  group_by(dow) %>%
  summarise(avg_load = mean(ERCOT_total_load_mw)) %>%
  ggplot(aes(dow, avg_load, group=1)) +
  geom_line() +
  labs(title="Average Load by Day of Week")

# Monthly pattern
df %>%
  group_by(month) %>%
  summarise(avg_load = mean(ERCOT_total_load_mw)) %>%
  ggplot(aes(month, avg_load, group=1)) +
  geom_line() +
  labs(title="Average Load by Month")
```

> Hourly: lowest aaround 4-5am, rises staring 6am, peaks around 5-6pm, drops after around 8pm.

> Weekly: lowest on Sunday, highest on Tue to Thu.

> Monthly: lowest in March, rapid increase from May to June, peak in August, drop in Oct and Nov


#### Autocorrelation

```{r}
Acf(df$ERCOT_total_load_mw, lag.max=24*14)
Pacf(df$ERCOT_total_load_mw, lag.max=24*14)
```
> ACF plot: The series is not stationary because ACF decays very slowly and stays high for many lags

```{r}
# difference the data
y_diff <- diff(df$ERCOT_total_load_mw, lag=1)
y_seasonal <- diff(y_diff, lag=24)

# plot
Acf(na.omit(y_seasonal), lag.max=200)
Pacf(na.omit(y_seasonal), lag.max=200)
```

#### Stationarity Test

```{r}
library(tseries)
adf.test(df$ERCOT_total_load_mw)

# seasonal differecing
adf.test(diff(df$ERCOT_total_load_mw, lag=24))

kpss.test(df$ERCOT_total_load_mw)

# First difference
kpss.test(diff(df$ERCOT_total_load_mw, lag=1))
```

> ADF says stationary and KPSS says non-stationary.

> Contradictory results: https://stats.stackexchange.com/questions/372145/adf-test-showing-stationary-for-a-non-stationary-series

> Explanation: The ADF test and KPSS test give different results because they have opposite null hypotheses and detect different types of non-stationarity. The ADF test tests for a unit root (stochastic trend), while the KPSS test tests for stationarity and is more sensitive to deterministic trend and seasonal structure. In our ERCOT load data, ADF rejects the unit root, but KPSS rejects stationarity, indicating the series is not a pure random walk but still contains strong seasonal persistence. This is why we apply both first differencing and seasonal differencing before fitting SARIMA models.


#### Weather Behavior (Zone Level)

```{r}
# hourly

library(tidyverse)
library(lubridate)

# Convert to long format
temp_long <- df %>%
  select(datetime,
         ends_with("_temperature_f")) %>%
  pivot_longer(-datetime,
               names_to = "zone",
               values_to = "temperature") %>%
  mutate(zone = str_remove(zone, "_temperature_f"))

# Faceted plot
ggplot(temp_long, aes(x = datetime, y = temperature)) +
  geom_line(alpha = 0.5) +
  facet_wrap(~ zone, ncol = 2) +
  labs(title = "Temperature Over Time by ERCOT Zone",
       y = "Temperature (F)",
       x = "Time") +
  theme_minimal()
```

> All 8 zones show strong annual seasonality and the seasonal shape is similar across zeons. Differences are mainly in peak summer temperatures and winter volatility.

```{r}
# daily
temp_daily <- temp_long %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date, zone) %>%
  summarise(temperature = mean(temperature), .groups = "drop")

ggplot(temp_daily, aes(x = date, y = temperature)) +
  geom_line() +
  facet_wrap(~ zone, ncol = 2) +
  labs(title = "Daily Average Temperature by ERCOT Zone",
       y = "Temperature (F)",
       x = "Date") +
  theme_minimal()
```

#### Load vs. Temperature

```{r}
library(tidyverse)

# Pivot temperature
temp_long <- df %>%
  select(datetime, ends_with("_temperature_f")) %>%
  pivot_longer(-datetime,
               names_to = "zone",
               values_to = "temperature") %>%
  mutate(zone = str_remove(zone, "_temperature_f"))

# Pivot load
load_long <- df %>%
  select(datetime, ends_with("_load_mw")) %>%
  pivot_longer(-datetime,
               names_to = "zone",
               values_to = "load") %>%
  mutate(zone = str_remove(zone, "_load_mw"))

# Merge them
combined <- left_join(temp_long, load_long,
                      by = c("datetime", "zone"))
```

```{r}
ggplot(combined,
       aes(x = temperature, y = load)) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  facet_wrap(~ zone, ncol = 2, scales = "free_y") +
  labs(title = "Load vs Temperature by ERCOT Zone",
       x = "Temperature (F)",
       y = "Load (MW)") +
  theme_minimal()
```

> U-shaped (nonlinear) relationship between load and temperature. Load is highest when temp is very low or high and lowest around 60-70F. Moreover, different zones respond differnetly to cold vs. heat, like FWEST has weaker temperature effect, suggesting zone-specific models.

#### Correlation Matrix

```{r}
library(dplyr)
library(corrplot)

zones <- c("COAST","EAST","FWEST","NORTH","NCENT","SOUTH","SCENT","WEST")

for (z in zones) {
  df_selected <- df %>%
    select(
      ERCOT_total_load_mw,
      all_of(paste0(z, "_temperature_f")),
      all_of(paste0(z, "_wind_speed_mph")),
      all_of(paste0(z, "_solar_radiation_wm2")),
      load_lag_24h,
      load_lag_168h
    )

  cor_matrix <- cor(df_selected, use = "complete.obs")

  corrplot(cor_matrix,
           method = "color",
           type = "full",
           addCoef.col = "black",
           number.cex = 0.7,
           tl.col = "black",
           tl.srt = 45,
           title = paste("Correlation Matrix:", z),
           mar = c(0, 0, 2, 0))   # gives room for title
}
```

> load_lag_24h: strong daily seasonality; load_lag_168h: weekly structure too; wind_speed: very weak relationship with load.

```{r}
library(dplyr)

load_df <- df %>%
  select(ends_with("_load_mw"))

cor_load <- cor(load_df, use = "complete.obs")

library(corrplot)

corrplot(cor_load,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         number.cex = 0.6,
         tl.col = "black",
         tl.srt = 45)
```

> FWEST load behaves more independently.

```{r}
temp_df <- df %>%
  select(ends_with("_temperature_f"))

cor_temp <- cor(temp_df, use = "complete.obs")

corrplot(cor_temp,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         number.cex = 0.6,
         tl.col = "black",
         tl.srt = 45)
```

> High temperature correlations.

#### Extreme Load Analysis

```{r}
threshold <- quantile(df$ERCOT_total_load_mw, 0.99)

extreme <- df %>%
  filter(ERCOT_total_load_mw >= threshold)

summary(extreme$COAST_temperature_f)
```

> Extreme demand in ERCOT is primarily driven by extreme heat. Some happen during very cold periods.

```{r}
# visualization of temperature

threshold <- quantile(df$ERCOT_total_load_mw, 0.99)

df$extreme_flag <- ifelse(df$ERCOT_total_load_mw >= threshold,
                          "Extreme", "Normal")

temp_long <- df %>%
  select(extreme_flag, ends_with("_temperature_f")) %>%
  pivot_longer(cols = -extreme_flag,
               names_to = "zone",
               values_to = "temperature")

ggplot(temp_long,
       aes(x = temperature, fill = extreme_flag)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ zone, scales = "free") +
  labs(title = "Temperature Distribution by Zone",
       x = "Temperature (F)")
```


```{r}
# heat index
heat_long <- df %>%
  select(extreme_flag, ends_with("_heat_index_f")) %>%
  pivot_longer(cols = -extreme_flag,
               names_to = "zone",
               values_to = "heat_index")

ggplot(heat_long,
       aes(x = heat_index, fill = extreme_flag)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ zone, scales = "free") +
  labs(title = "Heat Index Distribution by Zone")
```

```{r}
# relative humidity
rh_long <- df %>%
  select(extreme_flag, ends_with("_relative_humidity_pct")) %>%
  pivot_longer(cols = -extreme_flag,
               names_to = "zone",
               values_to = "humidity")

ggplot(rh_long,
       aes(x = humidity, fill = extreme_flag)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ zone, scales = "free") +
  labs(title = "Humidity Distribution by Zone")
```

> Extreme load tends to occur at lower-to-moderate humidity. The separation is weaker than temperature.

```{r}
# solar radiation
solar_long <- df %>%
  select(extreme_flag, ends_with("_solar_radiation_wm2")) %>%
  pivot_longer(cols = -extreme_flag,
               names_to = "zone",
               values_to = "solar")

ggplot(solar_long,
       aes(x = solar, fill = extreme_flag)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ zone, scales = "free") +
  labs(title = "Solar Radiation Distribution by Zone")
```

> Extreme load mostly happens when solar radiation is high, but many solar=0 observations (nighttime) dominate distribution

```{r}
# wind speed
wind_long <- df %>%
  select(extreme_flag, ends_with("_wind_speed_mph")) %>%
  pivot_longer(cols = -extreme_flag,
               names_to = "zone",
               values_to = "wind_speed")

ggplot(wind_long,
       aes(x = wind_speed, fill = extreme_flag)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ zone, scales = "free") +
  labs(title = "Wind Speed Distribution by Zone")
```

> No clear separation.

#### Weekend & Holiday Effects

```{r}
ggplot(df, aes(x = factor(is_weekend),
               y = ERCOT_total_load_mw)) +
  geom_boxplot() +
  labs(title = "Load: Weekend vs Weekday",
       x = "Is Weekend")

ggplot(df, aes(x = factor(is_holiday),
               y = ERCOT_total_load_mw)) +
  geom_boxplot() +
  labs(title = "Load: Holiday vs Non-Holiday",
       x = "Is Holiday")

table(df$extreme_flag, df$is_weekend)
table(df$extreme_flag, df$is_holiday)
```
> Extreme on weekend: 16%; Extreme on holiday: 1.7%

> Weekend median load is lower; extreme loads rarely happen on holidays.
